Windowed least-squares model reduction for dynamical systems

### 2019-10-10 cavity.tex
through conclusions

\item \textit{\methodAcronymROMs\ with collocation}: We consider \methodAcronymROMs\ solved via the direct method. The ROMs use the Crank--Nicolson time-discretization with a time step of 
$\Delta t = 0.1$. We deem the Gauss--Newton algorithm converged when the gradient norm is less than $1e-4$. All Jacobians are computed via automatic differentiation. Uniform quadrature 
weights are used for evaluating the integral in~\eqref{eq:obj_gen_slab}.

@@

\subsubsection{Numerical results}
We first assess the performance of \methodAcronymROMs\ with varying window sizes. To this end, we consider \methodAcronymROMs\ with uniform window 
sizes of $\Delta T^n \equiv \Delta T = 0.2,0.5,1.0,$ and $2.0$, along with the Galerkin and LSPG ROMs. Results are first considered for basis \#1 as described in 
Table~\ref{tab:rom_basis_details}. For all ROMs, we evolve 
the solution for $t \in [0,100]$. This comprises the same time interval used to construct the trial subspace. First, Figure~\ref{fig:cav_results1a} depicts the evolution of the pressure at the bottom wall in the midpoint of the computational domain, while Figure~\ref{fig:cav_results1b} depicts the evolution of the normalized $\elltwo$ state error of the various reduced-order models. Both the hyper-reduced Galerkin and LSPG ROMs blow up/fail to converge within the first several time units. The \methodAcronymROM\ minimizing the residual over a window of size $\Delta T = 0.2$ also fails to converge. The \methodAcronymROMs\ that minimize the residual over window sizes of $\Delta T \ge 0.5$ are seen to all be stable and accurate; the pressure response is well characterized and the normalized state errors are again less than $10\%$. The most notable discrepancy between the \methodAcronymROM\ and FOM solutions is a slight phase difference.

@@

\end{center}
\caption{Comparison of the pressure profiles obtained at the midpoint of the bottom wall (top) and normalized $\elltwo$ state errors (bottom) of various hyper-reduced ROMs to the full-order model solution.}
\label{fig:cav_results1}
\end{figure}

Figure~\ref{fig:cav_results2} shows the integrated error and objective function of the stable \methodAcronymROMs. It is seen that growing the window size over 
which the residual is minimized leads to a lower space--time residual, but not necessarily a lower $\elltwo$ error. This result is consistent with the previous numerical example. Next, Figure~\ref{fig:cav_wallclock}
shows the wall-clock times of the \methodAcronymROMs\ for $t \in [0,10]$ as compared to the LSPG ROM\footnote{Note that LSPG blew up at $t \approx 16.0$, so we focus on the first ten time units}. As expected, increasing the time-slab size again leads to an increase in computational cost. Minimizing the residual over a time-slab comprising 20 time-steps leads to a 2.5x increase in cost over LSPG.

\begin{figure}
\begin{center}
\begin{subfigure}[t]{0.45\textwidth}
\includegraphics[trim={0cm 0cm 0cm 0cm},clip,width=1.\linewidth]{figs/cavity/error_vs_window_basis2.pdf}
\caption{Integrated $\elltwo$ error}
\label{fig:cav_results2a}
\end{subfigure}
\begin{subfigure}[t]{0.45\textwidth}
\includegraphics[trim={0cm 0cm 0cm 0cm},clip,width=1.\linewidth]{figs/cavity/objective_vs_window_basis2.pdf}
\caption{Objective function} 
\label{fig:cav_results2b}
\end{subfigure}
\end{center}
\caption{Integrated error (left) and objective function (right) as a function of window size.}
\label{fig:cav_results2}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[trim={0cm 0cm 0cm 0cm},clip,width=0.49\linewidth]{figs/cavity/walltime_vs_window_compare_basis2.pdf}
\caption{Wall-clock times of \methodAcronymROMs\ with respect to the LSPG ROM}
\label{fig:cav_wallclock}
\end{center}
\end{figure}

@@

Next, we assess the performance of the various ROMs for basis \#2 as described in Table~\ref{tab:rom_basis_details}, which comprises a richer spatial basis.
 Figure~\ref{fig:cav_results1_basis1} shows the same pressure and error profiles as Figure~\ref{fig:cav_results1}, but for the enriched basis. The LSPG and Galerkin ROMs blow up faster as compared to Figure~\ref{fig:cav_results1}; LSPG fails to converge around $t \approx 8$ (opposed to $t \approx 16$), while Galerkin blows up almost immediately. The 
\methodAcronymROMs\ again yield improved performance: \methodAcronymROMs\ minimizing the residual over window sizes of $\Delta T \ge 0.5$ are seen to all be stable and accurate; the pressure response is well characterized and the normalized state errors are less than $5\%$. The \methodAcronymROMs\ with $\Delta T \ge 0.5$ leveraging the enriched basis yield more accurate results than \methodAcronymROMs\ using the coarse basis.

@@

\begin{subfigure}[t]{1.0\textwidth}
\includegraphics[trim={0cm 2.5cm 0cm 3cm},clip,width=1.\linewidth]{figs/cavity/error.pdf}
\caption{Normalized $\elltwo$ error}
\label{fig:cav_results1b_basis1}
\end{subfigure}
\end{center}
\caption{Comparison of the pressure profiles obtained at the midpoint of the bottom wall (top) and normalized $\elltwo$ state errors (bottom) of various hyper-reduced ROMs to the full-order model solution.}
\label{fig:cav_results1_basis1}
\end{figure}

@@

Finally, Figure~\ref{fig:cav_wallclock} shows the wall-clock times of the \methodAcronymROMs\ as compared to the LSPG ROMs for basis \#2. Increasing the window size again leads to an increase in computational cost. Minimizing the residual over a window comprising 20 time-steps leads to a 2.5x increase in cost over LSPG. This increase in computational cost is particularly reasonable in this example as LSPG yielded an unstable solution. 
\begin{figure}
\begin{center}
\includegraphics[trim={0cm 0cm 0cm 0cm},clip,width=0.49\linewidth]{figs/cavity/walltime_vs_window_compare.pdf}
\caption{Wall-clock times of \methodAcronymROMs\ with respect to the LSPG ROM.}
\label{fig:cav_wallclock}
\end{center}
\end{figure}

>>>>

\item \textit{\methodAcronymROMs\ with collocation}: We consider \methodAcronymROMs\ solved via the direct method. The ROMs use the Crank--Nicolson time-discretization with a time step of 
$\Delta t = 0.1$. We deem the Gauss--Newton algorithm converged when the gradient norm is less than $1e-4$. All Jacobians are computed via automatic differentiation. Uniform quadrature 
weights are used for evaluating the integral in the minimization problem~\eqref{eq:obj_gen_slab}.

@@

\subsubsection{Numerical results}
We first assess the performance of \methodAcronymROMs\ with varying window sizes. To this end, we consider \methodAcronymROMs\ with uniform window 
sizes of $\Delta T^n \equiv \Delta T = 0.2,0.5,1.0,$ and $2.0$, along with the Galerkin and LSPG ROMs. We first consider results for basis \#1 as described in 
Table~\ref{tab:rom_basis_details}. For all ROMs, we evolve 
the solution for $t \in [0,100]$. This comprises the same time interval used to construct the trial subspace. First, Figure~\ref{fig:cav_results1a} depicts the evolution of the pressure at the bottom wall in the midpoint of the computational domain, while Figure~\ref{fig:cav_results1b} depicts the evolution of the normalized $\elltwo$ state error of the various reduced-order models. Both the collocated Galerkin and LSPG ROMs blow up/fail to converge within the first several time units. The \methodAcronymROM\ minimizing the residual over a window of size $\Delta T = 0.2$ also fails to converge. The \methodAcronymROMs\ that minimize the residual over window sizes of $\Delta T \ge 0.5$ are seen to all be stable and accurate; the pressure response is well characterized and the normalized state errors are again less than $10\%$. The most notable discrepancy between the \methodAcronymROM\ and FOM solutions is a slight phase difference.

@@

\end{center}
\caption{Comparison of the pressure profiles obtained at the midpoint of the bottom wall (top) and normalized $\elltwo$ state errors (bottom) of various collocated ROMs to the full-order model solution.}
\label{fig:cav_results1}
\end{figure}

Figure~\ref{fig:cav_results2} shows the space--time error and objective function of the stable \methodAcronymROMs. It is seen that growing the window size over 
which the residual is minimized leads to a lower space--time residual, but not necessarily a lower $\elltwo$ error. This result is consistent with the previous numerical example. Next, Figure~\ref{fig:cav_wallclock}
shows the wall-clock times of the \methodAcronymROMs\ for $t \in [0,10]$ as compared to the LSPG ROM\footnote{Note that LSPG blew up at $t \approx 16.0$, so we focus on the first ten time units.}. As expected, increasing $\Delta T$ again leads to an increase in computational cost; minimizing the residual over a window comprising 20 time instances yields a 2.5x increase in cost over LSPG.

\begin{figure}
\begin{center}
\begin{subfigure}[t]{0.45\textwidth}
\includegraphics[trim={0cm 0cm 0cm 0cm},clip,width=1.\linewidth]{figs/cavity/error_vs_window_basis2.pdf}
\caption{Space--time $\elltwo$ error}
\label{fig:cav_results2a}
\end{subfigure}
\begin{subfigure}[t]{0.45\textwidth}
\includegraphics[trim={0cm 0cm 0cm 0cm},clip,width=1.\linewidth]{figs/cavity/objective_vs_window_basis2.pdf}
\caption{Objective function} 
\label{fig:cav_results2b}
\end{subfigure}
\end{center}
\caption{Space--time error (left) and objective function (right) as a function of window size.}
\label{fig:cav_results2}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[trim={0cm 0cm 0cm 0cm},clip,width=0.49\linewidth]{figs/cavity/walltime_vs_window_compare_basis2.pdf}
\caption{Wall-clock times of \methodAcronymROMs\ with respect to the LSPG ROM.}
\label{fig:cav_wallclock}
\end{center}
\end{figure}

@@

Next, we assess the performance of the various ROMs for basis \#2 as described in Table~\ref{tab:rom_basis_details}, which comprises a richer spatial basis.
 Figure~\ref{fig:cav_results1_basis1} shows the same pressure and error profiles as Figure~\ref{fig:cav_results1}, but for the enriched basis. The LSPG and Galerkin ROMs blow up faster as compared to Figure~\ref{fig:cav_results1}; LSPG fails to converge around $t \approx 8$ (opposed to $t \approx 16$), while Galerkin blows up almost immediately. The 
\methodAcronymROMs\ again yield improved performance: \methodAcronymROMs\ minimizing the residual over window sizes of $\Delta T \ge 0.5$ are seen to all be stable and accurate; the pressure response is well characterized and the normalized state errors are less than $5\%$. The \methodAcronymROMs\ with $\Delta T \ge 0.5$ and basis \#2 yield more accurate results than \methodAcronymROMs\ with basis \#1.

@@

\begin{subfigure}[t]{1.0\textwidth}
\includegraphics[trim={0cm 2.5cm 0cm 3cm},clip,width=1.\linewidth]{figs/cavity/error.pdf}
\caption{Normalized $\elltwo$ error}
\label{fig:cav_results1b_basis1}
\end{subfigure}
\end{center}
\caption{Comparison of the pressure profiles obtained at the midpoint of the bottom wall (top) and normalized $\elltwo$ state errors (bottom) of various collocated ROMs to the full-order model solution.}
\label{fig:cav_results1_basis1}
\end{figure}

@@

Finally, Figure~\ref{fig:cav_wallclock} shows the wall-clock times of the \methodAcronymROMs\ as compared to the LSPG ROMs for basis \#2. Increasing the window size again leads to an increase in computational cost. Minimizing the residual over a window comprising 20 time instances yields a 2.5x increase in cost over LSPG. This increase in computational cost is particularly reasonable in this example as LSPG yielded an unstable solution. 
\begin{figure}
\begin{center}
\includegraphics[trim={0cm 0cm 0cm 0cm},clip,width=0.49\linewidth]{figs/cavity/walltime_vs_window_compare.pdf}
\caption{Wall-clock times of \methodAcronymROMs\ with respect to the LSPG ROM.}
\label{fig:cav_wallclock}
\end{center}
\end{figure}

### 2019-10-10 main.tex
updated figure notation in cavity example
through conclusions

\section{Conclusions}\label{sec:conclude}
This paper introduced a windowed least-squares reduced-order modeling (\methodAcronym) formulation for dynamical systems. The approach sequentially minimizes the time-continuous full-order model residual within a low-dimensional trial subspace over a series of time windows. The formulation is applicable to both spatial and space--time dimension reduction trial subspaces. The Galerkin, Least-Squares Petrov--Galerkin (LSPG) and space--time LSPG methods were shown to be specific instances of the proposed method

Two differing techniques to solve the least-squares minimization problems were considered: direct (discretize then minimize) and indirect (minimize then discretize). 
The case of \spatialAcronym\ trial subspaces proved particularly interesting: it was shown that indirect methods comprise solving a coupled two-point 
Hamiltonian boundary value problem. The forward problem evolves the standard Galerkin ROM forced by a ``costate" variable, while a backward problem evolves the costate and is forced by the ROM residual.

Numerical experiments of the compressible Euler and Navier-Stokes equations demonstrated the utility in the proposed approach. The first numerical experiment, in where the Sod shock tube was examined, demonstrated that \methodAcronymROMs\ minimizing the residual over larger time windows yielded smoother and more physically relevant solutions. Further, increasing the window size over which the residual was minimized led to a monotonic decrease in the total space--time residual. Increasing the window size over which the residual was minimized, however, did not necessarily decrease the $\elltwo$ solution error; this was observed to occur over an intermediary window size. It was 
additionally shown that the \methodAcronymROMs\ overcome the time-step sensitivity challenges that LSPG is subject to. The second numerical experiment, which examined hyper-reduced ROMs of a compressible cavity flow, demonstrated the utility of the \methodAcronym\ formulation on a more complex flow. In this experiment, \methodAcronymROMs\ yielded predictions with relative errors less than $5-10\%$, while the Galerkin and LSPG ROMs failed to converge/blew up. Increasing the window size over which the residual was minimized again led to a lower space--time residual, but not necessarily a lower $\elltwo$ error.

@@

The principle challenge encountered in the \methodAcronym\ formulation is the computational cost: increasing the window size over which the residual is minimized leads to, in general, a higher computational cost. This increased cost is due to the fact that forming and solving the nonlinear least-squares problem becomes more expensive and difficult as the window size grows. While numerical experiments demonstrated that this increase in computational cost is mild, future work will focus on the development of solution techniques to improve the scaling of the method.

@@

where $\integrand: \RR{M} \otimes \timeSpaceArg{} \times \RR{M} \otimes \timeSpaceArg{} \times[0,T] \rightarrow  \RRplus$ (for arbitrary $M$) with $\integrand: (\stateyDiscrete,\stateyDotDiscrete,\timeDummy) \mapsto \integrand(\stateyDiscrete,\stateyDotDiscrete,\timeDummy)$. We now introduce the function $\statez: [0,T] \rightarrow \RR{M}$ that is implicitly defined as an extremum of~\eqref{eq:el_gen} subject to the boundary condition $\statez(0) = \statez_0$. 

>>>>

\section{Conclusions}\label{sec:conclude}
This paper introduced the windowed least-squares (\methodAcronym) \approachKwd\ for model reductio\  of dynamical systems. The approach sequentially minimizes the time-continuous full-order model residual within a low-dimensional trial subspace over a series of time windows. The formulation is applicable to both spatial and space--time dimension reduction trial subspaces. The Galerkin, Least-Squares Petrov--Galerkin (LSPG) and space--time LSPG methods were shown to be specific instances of the proposed method

We considered two differing techniques to solve the least-squares minimization problems arising from the \methodAcronym\ \approachKwd: direct (discretize-then-optimize) and indirect (optimize-then-discretize). 
The case of \spatialAcronym\ trial subspaces proved particularly interesting: it was shown that indirect methods comprise solving a coupled two-point 
Hamiltonian boundary value problem. The forward problem evolves the standard Galerkin ROM forced by a ``costate" variable, while a backward problem evolves the costate and is forced by the ROM residual.

Numerical experiments of the compressible Euler and Navier--Stokes equations demonstrated the utility in the proposed approach. The first numerical experiment, in where the Sod shock tube was examined, demonstrated that \methodAcronymROMs\ minimizing the residual over larger time windows yielded solutions with lower space--time residuals. Increasing the window size over which the residual was minimized, however, did not necessarily decrease the $\elltwo$ solution error; this was observed to occur over an intermediary window size. We additionally observed that the  
\methodAcronym\ \approachKwd\ overcomes the time-step sensitivity challenges that LSPG is subject to. The second numerical experiment, which examined collocated ROMs of a compressible cavity flow, demonstrated the utility of the \methodAcronym\ formulation on a more complex flow. In this experiment, \methodAcronymROMs\ yielded predictions with relative errors less than $5-10\%$, while the Galerkin and LSPG ROMs failed to converge/blew up. Increasing the window size over which the residual was minimized again led to a lower space--time residual, but not necessarily a lower $\elltwo$ error.

@@

The principle challenge encountered in the \methodAcronym\ formulation is the computational cost: increasing the window size over which the residual is minimized leads to, in general, a higher computational cost. In the context of the direct solution approach, this increased cost is due to the fact that forming and solving the least-squares problem becomes more expensive as the window size grows. In the context of indirect methods, this increased cost is a result of the increased number of iterations of the forward backward sweep method. While numerical experiments demonstrated that the increase in computational cost is mild, future work will focus on the development of solution techniques to improve the scaling of the method.

@@

where $\integrand: \RR{M}  \times \RR{M}  \times[0,T] \rightarrow  \RRplus$ (for arbitrary $M$) with $\integrand: (\stateyDiscrete,\stateyDotDiscrete,\timeDummy) \mapsto \integrand(\stateyDiscrete,\stateyDotDiscrete,\timeDummy)$. We now introduce the function $\statez: [0,T] \rightarrow \RR{M}$ that is implicitly defined as an extremum of~\eqref{eq:el_gen} subject to the boundary condition $\statez(0) = \statez_0$. 
